\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-01 doc/publications/plos16/plos16.tex
a few changes

% Template for PLoS
% Version 3.3 June 2016
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

%%local edits (rob)
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
{\raggedright\footnotesize #1}}
\reversemarginpar
\setlength{\marginparwidth}{5cm}

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\margin[ar{something wrong here. We do not report on chromosome 1 and 2} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
contitions casued by the interaction of large groups of SNPs. 
\marginpar{check this}

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a
``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a large
number of samples, this scalability eventuall break down as we include more features.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
homozygous variant (1\textbar1) as 2.

@@

\subsubsection{Synthetic data} 
\marginpar{these are not actually encoded as factors}
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trail with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance.
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
\marginpar{repicated methods section}
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB (out of bag)
classification error and the 20 most important variables.

\marginpar{the accuracy hardly changes}
Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

\marginpar{wrong figure cited}
Figure~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
%Piotr can you please explain the remaining error on the synthetic data
By $mtry=0.8$ the error is converging to X, which is due to Y.

@@

\begin{figure}[tbph]
\begin{adjustwidth}{-1.00in}{0in}
\begin{tabular}{ll}
a)& b)\\
\includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
\includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
\end{tabular}
\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for
  random forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
  importance of top 20 variabels normalized to the sum of the linear model coefficients. {\bf Others} represents
  aggregated importance of variables not present in the linear model.  {\bf b)} OOB (out of bag) classification error
  estimate as funtion of \mtry\ fraction.}
\label{figure:synth}
\end{adjustwidth}
\end{figure}

@@

In Table~\ref{table3}, we display the top 10 features, ranked by their Gini index in the random forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
\end{tabular}
\begin{flushleft} Here the importance is the gini coefficient when building a model using the cell type as the label.
\end{flushleft}
\label{table3}
%\end{adjustwidth}
\end{table}

@@

                                                    & ranger - R    & 0.94       & 4 1/2 mins       & 21GB (one processor) \\

@@

                                                    & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't

@@

                                                    & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the

@@

We investigate the different synthetic datasets generated for section 5.1. and measured the time taken to build a random
forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
same random seed, to improve consistency of measurements.

@@

\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
lower and higher values the performance degrades slightly (Fig 1 b). In the latter case the likely cause is
communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case its most likely caused by reaching the clusters computational
capacity.

@@

\begin{table}[!ht]
\caption{
{\bf Sensitivity analysis on parameter choice.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
\hline
&&&&&\\ \hline
\end{tabular}
\begin{flushleft} 
  Table notes here.
\end{flushleft}
\label{table2}
\end{table}

@@

>>>>

% Template for PLoS
% Version 3.3 June 2016
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example
% table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
%
% Below are a few tips to help format your equations and other special characters according to our specifications. For
% more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at
% http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example,
% x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as
% CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part
% of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change
% "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

%%local edits (rob)
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
{\raggedright\footnotesize #1}}
\reversemarginpar
\setlength{\marginparwidth}{5cm}
\usepackage{caption}
\usepackage{subcaption}

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\marginpar{something wrong here. We do not report on chromosome 1 and 2} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
conditions caused by the interaction of large groups of SNPs. 
\marginpar{check this}

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage disequilibrium?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a
``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a large
number of samples, this scalability eventually break down as we include more features.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite[]{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

\subsubsection{Synthetic data} 
\marginpar{these are not actually encoded as factors}
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to
  have a very poor
\marginpar{is this true}
  fit, and be useless for prediction, and still have a useful ranking of variable importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
\marginpar{replicated in methods section}
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.
\marginpar{in the terminology of statistics, these are all main effects. There are no interaction terms}

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the 20 most important variables.

\marginpar{the accuracy hardly changes}
Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

\marginpar{wrong figure cited}
Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
%Piotr can you please explain the remaining error on the synthetic data
By $mtry=0.8$ the error is converging to X, which is due to Y.

@@

\marginpar{figure needs bigger text}
\begin{figure}[tbhp] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
    \caption{Variable importance of top 20 variables.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=6cm]{./figs/var_oob.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
    forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
  importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
  aggregated importance of variables not present in the linear model. The dotted line represents the original linear
  model coefficients.
 {\bf b)} OOB (out of bag) classification error
  estimate as function of \mtry\ fraction.}
  \label{figure:synth}
\end{figure}

@@

In Table~\ref{table3}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.
\marginpar{This is the Gini impurity measure, The Gini index or coefficient is an economic measure of income inequality}

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
\end{tabular}
\begin{flushleft} Here the importance is the Gini impurity measure when building a model using the cell type as the label.
\end{flushleft}
\label{table3}
%\end{adjustwidth}
\end{table}

@@

                                                    & ranger - R    & 0.06       & 4 1/2 mins       & 21GB (one processor) \\

@@

                                                    & ranger - R      &  0.05       & 28 mins          & num.threads=10\footnote{don't

@@

                                                    & ranger - R      & 0.03      & 58 mins\footnote{ but took 6 hours to load the

@@

We investigate the different synthetic datasets generated for section 5.1. and measured the time taken to build a random
forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
same random seed, to improve the consistency of measurements.

@@

\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
lower and higher values the performance degrades slightly (Fig 1 b). In the latter case the likely cause is
communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters computational
capacity.

@@

% \begin{table}[!ht]
% \caption{
% {\bf Sensitivity analysis on parameter choice.}}
% \begin{tabular}{|l|l|l|l|l|l|l|l|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{ \mtry\ }  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
% \hline
% &&&&&\\ \hline
% \end{tabular}
% \begin{flushleft} 
%   Table notes here.
% \end{flushleft}
% \label{table2}
% \end{table}

@@

% \begin{figure}[tbph]
% \begin{adjustwidth}{-1.00in}{0in}
% \begin{tabular}{ll}
% a)& b)\\
% \includegraphics[totalheight=6cm]{./figs/var_oder.png} & 
% \includegraphics[totalheight=6cm]{./figs/var_oob.png} \\
% \end{tabular}
% \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
%     forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
%   importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
%   aggregated importance of variables not present in the linear model. The dotted line represents the original linear
%   model coefficients.
%  {\bf b)} OOB classification error
%   estimate as function of \mtry\ fraction.}
% \label{figure:synth}
% \end{adjustwidth}
% \end{figure}
